Practical Challenges with Web Service Evolution:\\A Review of Literature and Current Industry Practices

### 2013-06-03 paper.tex
reorganized the contents in more of compare-contrast fashion

In this paper we identify the key challenges with web service evolution such as backward and forward compatibility, serving and testing multiple concurrent versions, version specific documentation, communicating future changes, etc. Following an appropriate strategy to overcome these challenges will enable web service publishers to iterate their services as required while minimizing the impact on existing consumers of the services. In order to find an appropriate strategy, the web service publisher needs to balance a tradeoff between the business drivers and technical challenges. To understand and deal with these challenges, we performed a systematic review of the existing literature on web service evolution and investigated some real world web services. Based on our findings, we discuss the challenges and suggested pragmatic techniques to address the challenges.

\keywords{Web Service, Evolution, Versioning, Adapt, SOA, SOAP, REST}
\end{abstract}

@@

@@

The term Representational State Transfer (REST) was introduced by Roy Fielding \cite{rest_wiki} and is used interchangeably as RESTful web service or Web API. In brief, REST is an implementation for web services where user defined ``resources'' are created, accessed, modified and deleted utilizing the various features of the HTTP stack such as caching, security, layering and different request methods (GET, POST, PUT, DELETE etc.), content types, etc.

Due to their historical usage, the term ``web service'' is often interpreted as SOAP. However, throughout this paper, unless otherwise specified, we have used this term to represent a web based communication method between two software systems irrespective of their actual implementation details. The findings mentioned in this paper are based on SOAP and REST based services due to their relative popularity over the other types.
% subsection restful_web_services (end)

@@

\section{Motivation} % (fold)
\label{sec:motivation}
Web services, like other software systems, need to evolve to stay on top of the bugs and ever changing requirements. For example, between May 15th 2013 and May 22nd 2013, the Facebook API team saw 238 new bugs, resolved 211 bugs and fixed 24 bugs \cite{facebook_release_note}. However, often times web services are consumed by an unknown number of consumers, and the web service publishers have little control over them. Updating a web service can be tricky in such a situation since a breaking change may upset the existing consumers. To be able to evolve in such a decoupled scenario, multiple versions of a service need to coexist which poses both technical and business challenges in multiple fronts. At a conceptual level, Figure \ref{fig:web_service_layers} shows the different aspects of a web service that need to deal with evolution challenges:

\begin{figure*}[ht]
  \centering
  \includegraphics[width=\textwidth]{web_service_layers.png}
    \caption{Different Aspects of an Evolving Web Service}
  \label{fig:web_service_layers}
\end{figure*}

However, there is a little to no standard approach to tackle these aspects of an evolving web service. The motivation for this research stems from the need to improve this situation.

@@

To address the aforementioned research questions, we have limited our search to two high level sources as follows:

@@

We have performed a Systematic Literature Review to extract information from the published articles following \cite{kitchenham2007guidelines}. We have also investigated three published real-word web services to find out the answers to the research questions based on todays solutions.

@@

@@

\subsection{Literature Review} % (fold)
\label{sub:literature_review}

@@

One or more of these roles may be played by one individual. The stakeholders take part in driving changes to a web service. The authors identified the following types of changes in a web services: \textbf{Interface, Implementation, QoS, Usage, Requirement, SLA, Pre and post conditions, and Feedback}.

A change in any of the above impacts one or more of the stakeholder roles based on its type. In addition to stakeholder impacts, the sources are also interdependent on each other. For example, when a web service \textbf{provider} decides to change the \textbf{interface}, the \textbf{developer} needs to write the changed interface and its \textbf{implementation}. If the changed interface if not backward compatible, and the service \textbf{integrator} needs to modify the integration. The new interface and implementation may instigate a change on it’s \textbf{QoS, Usage and Pre-post conditions}, which may impact the \textbf{end user}.

@@

@@

\subsubsection{How to evolve the web services?} % (fold)
\label{sub:how_to_evolve_the_web_services_}
Now that we have discussed the drivers behind web service evolution, next we present the different approaches to tackle web service evolution based on the existing literature.

\subsubsection{Trade-offs:} % (fold)
Lublinsky compared between multiple approaches of implementing evolving web services based on different dimensions as follows \cite{lublinsky2007versioning}.

@@

  \item \textbf{Units of versioning}: One approach to version a service is to apply a single version to all it’s methods that can be deployed as a group. Another approach is to version individual methods or operations of the service as they change, this confines the change to only specific methods and provides a finer grained versioning. However, deployment of such versioned methods are complicated and it forces the consumer to specify the service, its method name and an additional version to use for each operation.
  \item \textbf{Service changes, constituting a new version}: In line with \cite{treiber2009analyzing}  Lublinsky also pointed out that a new version may be required primarily because of changes in its interface, schema of the message payload or in it’s implementation.
  \item \textbf{Service version life-cycle considerations}: Based on the actual need, Lublinsky suggests adjusting the life-cycle of versions, as too many versions increase the maintenance overhead while too few versions leave the consumers a smaller time window to upgrade.
  \item \textbf{Version deployment/access approaches}: When deploying versioned services, Lublinsky compares between multiple versions deployed on the same endpoint vs. each version deployed on its own endpoint. Lublinsky points out that the former approach may introduce conflicts in names of the objects, databases and other related resources, and requires intermediate routers, which typically lowers performance and may introduce SLA complications. The author favored the latter approach and argued that it provides better scalability and loose coupling between versions.

@@

\subsubsection{Versioning:}
When multiple versions of a service are deployed at the same time, a versioning scheme is required to identify and use a specific version. Leskey discusses about the versioning of web services in \cite{laskey2008considerations} with respect to the OASIS SOA Reference Model (SOA-RM). In this paper, the author discussed versions in terms of identifiable resources, that can have multiple versions and are modified by applying new versions to the original one.

@@

The author then identified the key reasons for changes between versions such as: a) service description changes, b) changes in functions, c) changes in the interaction mechanics, and d) change of known pre and post conditions. These changes would require new versions of the service as well as accompanying any explanation of the specific changes.

@@

\subsection{Review Of Industry Examples} % (fold)
\label{sub:review_of_industry_examples}

To answer the research questions, in addition to looking at the published articles, we have also reviewed the following three popular web services from the industry:

A comprehensive review of all the real world web services is beyond the scope of this research. However, these selected web services are evolving with a large number of existing consumers using different approaches to evolve as described next.

\subsubsection{Google Maps:}
Google Maps provides a JavaScript based web service for it’s consumers and updates the service to provide ``new features, bug fixes and performance improvements'' \cite{google_maps_versioning}. Their versioning policy offers the following versions:

\begin{itemize}
  \item \textbf{Experimental version}: With all the latest features, bug fixes and improvements. But this version is not guaranteed to be feature stable.
  \item \textbf{Release version}: From the the experimental version, every quarter, a feature stable version of the service is tagged as a ``release'' version. These release versions may receive bug fixes but the features remain stable.
  \item \textbf{Previously released or Frozen version}: Every time a new ``release version'' is tagged, the oldest existing release version is ``frozen'', so no new features or bug fixes are applied to ensure it remains completely feature stable.
\end{itemize}

The consumers can specify the desired version using a version parameter in the URL. For example, to use the version 3.12, the following URL can be used https://maps.googleapis.com/maps/api/js?v=3.12 In absence of a version parameter, by default the experimental version is used.
Every time a new release version is tagged, the oldest version is retired. Google Maps guarantees the versions are backward compatible and automatically migrates the consumers using the oldest version to the following version without needing the consumers to make any change on their ends.

The Google Maps API has HTML based documentations and changelogs for each version. To communicate changes with the consumers, an email groupware is used.

\subsubsection{Twitter API:}
Twitter provides a RESTful web service to interact with it’s data from third-party applications \cite{twitter_api}. This web service is also evolving to fix bugs and bring new features, but using a different approach from that of Google Maps.

The consumers can specify the target version in the URL. For example, the web service calls to https://api.twitter.com/1.1/ will use version 1.1.

Once a version is released, future changes are made to the released version to provide new features and bug fixes. However, unlike Google Maps, new features are not always released under a new version name. Instead, the changes may be deployed under the same version name and are communicated to the consumers using an updated “Calendar of API Changes” \cite{twitter_calendar} through a developer portal.

The developer portal is also used to record issues and collect feedback from the consumers. The portal contains version specific HTML based documentations of the web service as well as an interactive web browser based console to explore the service on the browser \cite{twitter_console} .

\subsubsection{Facebook Graph API:}

Unlike the previous approaches, Facebook Graph API on the other hand uses a different approach for it’s evolution. Facebook Graph API is the primary web service to retrieve and post data to Facebook \cite{facebook_api}. They have a 90-day breaking change policy stated as follows:

@@

They also identified some example of the change drivers such as, change/removal of major functionality, backwards compatibility, Facebook product changes, and privacy and security related changes. For each new version, instead of a URL based addressing scheme as used by Google Maps and Twitter API, Facebook allows the consumers to enable the changes by pushing a button on the Facebook website. Using the button, the service consumers, aka platform developers, can opt-in to migrate to a new release to get an early look on the impact of the breaking changes specific to their use case and take necessary actions within the 90-day period.

To communicate the web service changes with it’s consumers and collect feedback, Facebook uses a developer portal as well as a StackOverflow channel \cite{facebook_stack_overflow}. The developer portal shares information about planned upcoming changes as well as the release notes of each version.

The Facebook Graph API accompanies both a static HTML based and an interactive documentation of the API called ``Graph API Explorer''. As the name suggests, the interactive API explorer lets a consumer to explore the features of the service using live data on the browser without needing to write any code.

% subsection review_of_industry_examples (end)

>>>>

Service-oriented Systems often rely on integrating software systems that are developed by independent parties and that evolve independently. In this paper we identify the key challenges with web service evolution such as backward and forward compatibility, serving and testing multiple concurrent versions, version specific documentation, communicating future changes. Following an appropriate strategy to overcome these challenges will enable web service publishers to iterate their services as required while minimizing the impact on existing consumers of the services. To understand and deal with these challenges, we performed a systematic review of the existing literature on web service evolution and investigated three real world web services. Based on our findings, we discussed the challenges and opportunities for future work to address the challenges.

\keywords{Web Service, Evolution, Versioning, SOA, SOAP, REST}
\end{abstract}

@@

Due to their historical usage, the term ``web service'' is often interpreted as SOAP. However, throughout this paper, unless otherwise specified, we have used this term to represent a web based communication method between two software systems irrespective of their actual implementation details. The findings mentioned in this paper are based on SOAP and REST based services due to their relative popularity over the other types.

@@

The term Representational State Transfer (REST) was introduced by Roy Fielding \cite{rest_wiki} and is used interchangeably as RESTful web service or Web API. In brief, REST is an implementation for web services where user defined ``resources'' are created, accessed, modified and deleted utilizing the various features of the HTTP stack such as caching, security, layering and different request methods (GET, POST, PUT, DELETE etc.), content types.

% subsection restful_web_services (end)

@@

\section{Motivation} % (fold)
\label{sec:motivation}
Web services, like other software systems, need to evolve to stay on top of bugs and ever changing requirements. For example, between May 15th 2013 and May 22nd 2013, the Facebook API team saw 238 new bugs, resolved 211 bugs and fixed 24 bugs \cite{facebook_release_note}. However, often times web services are used by an unknown number of consumers, and the web service publishers have little control over them. Updating a web service can be tricky in such a situation since a breaking change may upset the existing consumers. To be able to evolve in such a decoupled scenario, multiple versions of a service need to coexist which poses both technical and business challenges in multiple fronts. At a conceptual level, Figure \ref{fig:web_service_layers} shows the different parts of a web service that need to deal with evolution challenges:

\begin{figure*}[ht]
  \centering
  \includegraphics[width=\textwidth]{web_service_layers.png}
    \caption{Different Parts of an Evolving Web Service}
  \label{fig:web_service_layers}
\end{figure*}

As shown in Figure \ref{fig:web_service_layers}, the different parts of a web service need to cope with the changing requirements for its evolution. However, there is a little to no standard approach to tackle these parts of an evolving web service. The motivation for this research stems from the need to improve this situation.

@@

We have performed a Systematic Literature Review to extract information from the published articles following \cite{kitchenham2007guidelines}. We have also investigated three published real-word web services to find out the answers to the research questions based on todays solutions.
In this regard, we have limited our research to two high level sources as follows:

@@

@@

\subsection{Analysis of Real World Web Service Evolution} % (fold)
\label{sub:analysis_of_real_world_web_service_evolution}
In addition to looking at the published articles, we have also analyzed real world web services to examine their approaches to deal with evolution. A comprehensive review of all the real world web services is beyond the scope of this research. However, we selected the following web services as they are evolving with a large number of existing consumers using different evolution approaches:

The triangulation of the findings from the systematic literature review and the analysis of real world services provide a broader insight than each method itself.

% subsection analysis_of_real_world_web_service_evolution (end)

@@

\subsection{Systematic Literature Review} % (fold)
\label{sub:literature_review}

@@

One or more of these roles may be played by one individual. The stakeholders take part in driving changes to a web service. The authors identified the following types of changes in a web services: Interface, Implementation, Quality of Service (QoS), Usage, Requirement, Service-level Agreement (SLA), Pre and post conditions, and Feedback. In line with \cite{treiber2009analyzing}  Lublinsky also pointed out that a new version may be required primarily because of changes in its interface, schema of the message payload or in it’s implementation. In \cite{laskey2008considerations}, Laskey identified the key reasons for changes between versions as: a) service description changes, b) changes in functions, c) changes in the interaction mechanics, and d) change of known pre and post conditions. These changes would require new versions of the service as well as accompanying any explanation of the specific changes.

@@

A change in a web service impacts one or more of the stakeholder roles based on its type. Treiber et al. pointed out that, in addition to stakeholder impacts, the sources are also interdependent on each other \cite{treiber2009analyzing}. For example, when a web service \textbf{provider} decides to change the \textbf{interface}, the \textbf{developer} needs to write the changed interface and its \textbf{implementation}. If the changed interface if not backward compatible, and the service \textbf{integrator} needs to modify the integration. The new interface and implementation may instigate a change its \textbf{QoS}, since it may improve or degrade the performance. The changes may improve or degrade the \textbf{QoS} and impact the \textbf{end user}.

@@

\subsubsection{How to evolve the web services?} % (fold)
\label{sub:how_to_evolve_the_web_services_}
Now that we have listed the drivers behind web service evolution, next we present the different approaches to tackle web service evolution based on the existing literature.

\subsubsection{Versioning:} % (fold)
Lublinsky compared between multiple approaches of implementing evolving web services based on different dimensions as follows \cite{lublinsky2007versioning}.

@@

  \item \textbf{Units of versioning}: One approach to version a service is to apply a single version to all it’s methods that can be deployed as a group. Another approach is to version individual methods or operations of the service as they change, this confines the change to only specific methods and provides a finer grained versioning. However, deployment of such versioned methods are complicated and it forces the consumer to specify the service, its method name and an additional version to use for each operation.
  \item \textbf{Service version life-cycle considerations}: Based on the actual need, Lublinsky suggests adjusting the life-cycle of versions, as too many versions increase the maintenance overhead while too few versions leave the consumers a smaller time window to upgrade since each version has a shorter life-cycle.
  \item \textbf{Version deployment/access approaches}: When deploying versioned services, Lublinsky compares between multiple versions deployed on the same endpoint vs. each version deployed on its own endpoint. Lublinsky points out that the former approach may introduce conflicts in names of the objects, databases and other related resources, and requires intermediate routers, which typically lowers performance and may introduce SLA complications. The author favored the latter approach and argued that it provides better scalability and loose coupling between versions.

@@

A versioning scheme is required to identify and use a specific version. Laskey discusses about the versioning of web services in \cite{laskey2008considerations}. In this paper, the author discussed versions in terms of identifiable resources, that can have multiple versions and are modified by applying new versions to the original one.

@@

@@

\subsection{Analysis of Real World Web Service Evolution} % (fold)
\label{sub:review_of_industry_examples}

In this section we have listed our findings from the following real world evolving web services:

\begin{itemize}
  \item \textbf{Google Maps}: Google Maps provides a web service for its consumers and updates the service to provide ``new features, bug fixes and performance improvements'' \cite{google_maps_versioning}.
  \item \textbf{Twitter Web API}: Twitter provides a RESTful web service to interact with its data from third-party applications \cite{twitter_api}. This web service is also evolving to fix bugs and bring new features, but using a different approach from that of Google Maps.
  \item \textbf{Facebook Graph API}: Facebook Graph API is the primary web service to retrieve and post data to Facebook \cite{facebook_api}. The service evolves to support change/removal of functionality, backwards compatibility, Facebook product changes, and privacy and security related changes.
\end{itemize}

\subsubsection{Versioning:} % (fold)
\label{sub:versioning}
Google Maps uses the following versioning policy \cite{google_maps_versioning}:

\begin{itemize}
  \item \textbf{Experimental version}: With all the latest features, bug fixes and improvements. But this version is not guaranteed to be feature stable.
  \item \textbf{Release version}: From the the experimental version, every quarter, a feature stable version of the service is tagged as a ``release'' version. These release versions may receive bug fixes but the features remain stable.
  \item \textbf{Previously released or Frozen version}: Every time a new ``release version'' is tagged, the oldest existing release version is ``frozen'', so no new features or bug fixes are applied to ensure it remains completely feature stable.
\end{itemize}
Every time a new release version is tagged, the oldest version is retired. Google Maps guarantees the versions are backward compatible and automatically migrates the consumers using the oldest version to the following version without needing the consumers to make any change on their ends.

On the other hand, the Twitter API had two versions at the time of writing this paper \cite{twitter_api} and the old version is marked as deprecated. The users of the old version are suggested to upgrade to the new version before its posted date of expiry. Once a version is released, future changes are made to the released version to provide new features and bug fixes. However, unlike Google Maps, new features are not always released under a new version name. Instead, the changes may be deployed under the same version name.

Unlike the previous approaches, Facebook Graph API on the other hand uses a different approach for its evolution. They have a 90-day breaking change policy stated as follows: \cite{facebook_api}

@@

The service consumers, aka platform developers, can opt-in to migrate to a new release to get an early look on the impact of the breaking changes specific to their use case and take necessary actions within the 90-day period. All users are automatically migrated to the new version after this 90-day period.

Both Google Maps and Twitter web service users identify a desired version of the service in the URL. For example, to use the version 3.12 of Google Maps API, the following URL can be used https://maps.googleapis.com/maps/api/js?v=3.12. In absence of a version parameter, by default the experimental version of Google Maps is used.

Instead of a URL based versioning identifier, Facebook allows the consumers to enable the changes by pushing a button on the Facebook website. Once enabled, the users can make necessary changes before the 90-day period.

\subsubsection{Documentation:}
Google Maps web service provides a human readable documentation of the service operations with example requests and responses specific to a version \cite{google_maps_services}. Since the API is publicly available, the users can run the example requests to produce live results by browsing the API endpoints on their browsers.

Twitter also publishes human readable API documentation for each version of the web service in their developer portal \cite{twitter_api}. The portal also includes an interactive web browser based console to let its users explore the service on the browser \cite{twitter_console}.

Similarly, the Facebook Graph API accompanies both a static HTML based and an interactive the API called ``Graph API Explorer'' \cite{facebook_api}.

\subsubsection{Communication:}

The real-world web services use multiple channels to communicate with the users. For example, Google Maps web service uses StackOverflow.com to collect code related questions from the users, an issue tracker to let the users report defects, and blogs and an email group to discuss changes \cite{google_maps_forum}.

To communicate the past and upcoming web service changes, Twitter publishes a ``Calendar of API Changes'' on their developer portal \cite{twitter_calendar}. In addition to this calendar, the portal also has developer blogs, a discussion forum and an issue tracker for its users.

Facebook also uses a developer portal, a StackOverflow channel and blogs to communicate with its users \cite{facebook_stack_overflow}. Among other things, the developer portal announces upcoming changes and effective dates of the 90-day breaking changes.
% subsection review_of_industry_examples (end)
